---
title       : EM clustering 
subtitle    : Example of Old Faitful data / Data Products assignment
author      : Dusan Randjelovic
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## EM clustering

Recently, I found out about expectation-maximisation algorithm, and was fascinated with how easily it could be understood. This project is a demo of clustering data with this algorithm.

Data used is Old Faitful geyser eruption data. When initial variables are set, algorithm converges in set number of iterations.

EM here was done using `mvtnorm` library, to fit data to multivariate normal distributions.

Final shiny app could be found [HERE](https://duxan.shinyapps.io/Data_products_EM_example/)

--- 

**Initialising**: User should first provide initialising data for both clusters, which will produce (reload) first plot, like here too

```{r echo=FALSE}
  #load library for multivariate normal
  library(mvtnorm)

  #load Old Faithful data frame
  data(faithful)

  #setup grid for plotting
  xpts <- seq(from=1,to=6,length.out=100)
  ypts <- seq(from=40,to=100,length.out=100)

  #initial parameter setup (chosen to be deliberately bad)
  theta <- list(
      tau=c(0.5,0.5),
      mu1 = c(4, 40), 
      mu2 = c(2, 50), 
      sigma1=matrix(c(0.8,7,7,70),ncol=2),
      sigma2=matrix(c(0.8,7,7,70),ncol=2)
    )

  #E step: calculates conditional probabilities for latent variables
  E.step <- function(theta){
    t(apply(cbind(
      theta$tau[1] * dmvnorm(faithful,mean=theta$mu1,sigma=theta$sigma1),
      theta$tau[2] * dmvnorm(faithful,mean=theta$mu2,sigma=theta$sigma2)
    ),1,function(x) x/sum(x)))
  }

  #M step: calculates the parameter estimates which maximise Q
  M.step <- function(T){
    list(
      tau= apply(T,2,mean),
      mu1= apply(faithful,2,weighted.mean,T[,1]),
      mu2= apply(faithful,2,weighted.mean,T[,2]),
      sigma1= cov.wt(faithful,T[,1])$cov,
      sigma2= cov.wt(faithful,T[,2])$cov
    )
  }

  #function to plot current data
  plot.em.initial <- function(theta){
    mixture.contour <- outer(xpts,ypts,function(x,y) {
      theta$tau[1]*dmvnorm(cbind(x,y),mean=theta$mu1,sigma=theta$sigma1) + theta$tau[2]*dmvnorm(cbind(x,y),mean=theta$mu2,sigma=theta$sigma2)
    })
    contour(xpts,ypts,mixture.contour,nlevels=5,drawlabel=FALSE,col="red",xlab="Eruption time (mins)",ylab="Waiting time (mins)",main=expression("Old Faithful data w/ initial params"))
    points(faithful)
  }
  
    # run code
    T <- E.step(theta)
    theta <- M.step(T)
    plot.em.initial(theta)

```

---

**Iterations**: After that user can set how many iterations algorithm should run, which updates second plot, like here too (27 iterations set, for previous initialising data)

```{r echo=FALSE}

   #initial parameter setup (chosen to be deliberately bad)
  iter <- 27
  theta.iter <- list(
      tau=c(0.5,0.5),
      mu1=c(2.8,75),
      mu2=c(3.6,58),
      sigma1=matrix(c(0.8,7,7,70),ncol=2),
      sigma2=matrix(c(0.8,7,7,70),ncol=2)
    )
  
  #E step: calculates conditional probabilities for latent variables
  E.step <- function(theta){
    t(apply(cbind(
      theta$tau[1] * dmvnorm(faithful,mean=theta$mu1,sigma=theta$sigma1),
      theta$tau[2] * dmvnorm(faithful,mean=theta$mu2,sigma=theta$sigma2)
    ),1,function(x) x/sum(x)))
  }

  #M step: calculates the parameter estimates which maximise Q
  M.step <- function(T){
    list(
      tau= apply(T,2,mean),
      mu1= apply(faithful,2,weighted.mean,T[,1]),
      mu2= apply(faithful,2,weighted.mean,T[,2]),
      sigma1= cov.wt(faithful,T[,1])$cov,
      sigma2= cov.wt(faithful,T[,2])$cov
    )
  }

  #function to plot data after iterations
  plot.em.iter <- function(theta){
    mixture.contour <- outer(xpts,ypts,function(x,y) {
      theta$tau[1]*dmvnorm(cbind(x,y),mean=theta$mu1,sigma=theta$sigma1) + theta$tau[2]*dmvnorm(cbind(x,y),mean=theta$mu2,sigma=theta$sigma2)
    })
    contour(xpts,ypts,mixture.contour,nlevels=5,drawlabel=FALSE,col="red",xlab="Eruption time (mins)",ylab="Waiting time (mins)",main=bquote("Old Faithful data after" ~ .(iter) ~ "iteration(s)"))
    points(faithful)
  }
  
    # run code
    for (i in 1:iter){
      T <- E.step(theta.iter)
      theta.iter <- M.step(T)
    }
    plot.em.iter(theta.iter)

```

---

Shiny app: [duxan.shinyapps.io](https://duxan.shinyapps.io/Data_products_EM_example/)  

THIS CODE GITHUB: [github.com/duxan](https://github.com/duxan/data-products-assignment-slidify)

## Thank you!

